\thetitle

### 2025-05-07 workshop/SPI.tex
almost 4 pages but needs work

To avoid the issue of non-associativity of 
floating point arithmetic, we ensure that the order in which operations are performed 
is exactly the same, irrespective of the number of processes/machines and threads. 
This is achieved by making sure that every value to be added---if addition is our 
reduction operation---is a leaf node in a reduction tree that is invariant to the 
number of machines available to perform the computation. 
For instance, if $N$ values are to be reduced, then the reduction tree would have 
$N$ leaf nodes. If $M$ machines are available, these machines are then assigned 
in such a way that the order of operations is as if there were $N$ machines available. 
To do so, it may be necessary for a machine to ``communicate with itself'', 
imitating the behaviour that would be present if there were $N$ machines available.
\cref{fig:reduction_tree_colour_8} and \cref{fig:reduction_tree_colour_2} 
illustrate the reduction procedure for eight and two machines, respectively.

@@

\begin{figure}[t]
  \centering
  \begin{forest}
    for tree = {circle, draw, minimum size = 0.8cm}
      [{$36x$}, fill = julia1
        [{$10x$}, fill = julia1
          [{$3x$}, fill = julia1
            [{$1x$}, fill = julia1]
            [{$2x$}, fill = julia2]
          ]
          [{$7x$}, fill = julia3
            [{$3x$}, fill = julia3] 
            [{$4x$}, fill = julia4] 
          ]
        ]
        [{$26x$}, fill = julia5
          [{$11x$}, fill = julia5
            [{$5x$}, fill = julia5] 
            [{$6x$}, fill = julia6] 
          ]
          [{$15x$}, fill = julia7
            [{$7x$}, fill = julia7] 
            [{$8x$}, fill = julia8] 
          ]
        ]
      ]
  \end{forest}
  \caption{Addition of eight floating point numbers across eight machines with a 
  guarantee on SPI. Each machine is represented by a different colour. 
  The final result can be represented 
  by the expression $((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x))$.}
  \label{fig:reduction_tree_colour_8}
\end{figure}

\begin{figure}[t]
  \centering
  \begin{forest}
    for tree = {circle, draw, minimum size = 0.8cm}
      [{$36x$}, fill = julia1
        [{$10x$}, fill = julia1
          [{$3x$}, fill = julia1
            [{$1x$}, fill = julia1]
            [{$2x$}, fill = julia1]
          ]
          [{$7x$}, fill = julia1
            [{$3x$}, fill = julia1] 
            [{$4x$}, fill = julia1] 
          ]
        ]
        [{$26x$}, fill = julia2
          [{$11x$}, fill = julia2
            [{$5x$}, fill = julia2] 
            [{$6x$}, fill = julia2] 
          ]
          [{$15x$}, fill = julia2
            [{$7x$}, fill = julia2] 
            [{$8x$}, fill = julia2] 
          ]
        ]
      ]
  \end{forest}
  \caption{Addition of eight floating point numbers across two machines with a 
  guarantee on SPI. Each machine is represented by a different colour. 
  The final result can be represented 
  by the expression $((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x))$,
  which is the same as that given by eight machines in \cref{fig:reduction_tree_colour_8}.}
  \label{fig:reduction_tree_colour_2}
\end{figure}

>>>>

To avoid the issue of non-associativity of 
floating point arithmetic, we ensure that the order in which operations are performed 
is exactly the same, irrespective of the number of processes/machines and threads. 
This is achieved by making sure that every value to be added---if addition is our 
reduction operation---is a leaf node in a reduction tree that is invariant to the 
number of machines available to perform the computation. 
For instance, if $N$ values are to be reduced, then the reduction tree would have 
$N$ leaf nodes. If $M$ machines are available, these machines are then assigned 
in such a way that the order of operations is as if there were $N$ machines available. 
To do so, it may be necessary for a machine to ``communicate with itself'', 
imitating the behaviour that would be present if there were $N$ machines available.
\cref{fig:reduction_tree_colour} illustrates the reduction procedure for eight and 
two machines.

@@

\begin{figure*}[t]
  \centering
  \begin{minipage}{0.5\linewidth}
  \begin{forest}
    for tree = {circle, draw, minimum size = 0.8cm}
      [{$36x$}, fill = julia1
        [{$10x$}, fill = julia1
          [{$3x$}, fill = julia1
            [{$1x$}, fill = julia1]
            [{$2x$}, fill = julia2]
          ]
          [{$7x$}, fill = julia3
            [{$3x$}, fill = julia3] 
            [{$4x$}, fill = julia4] 
          ]
        ]
        [{$26x$}, fill = julia5
          [{$11x$}, fill = julia5
            [{$5x$}, fill = julia5] 
            [{$6x$}, fill = julia6] 
          ]
          [{$15x$}, fill = julia7
            [{$7x$}, fill = julia7] 
            [{$8x$}, fill = julia8] 
          ]
        ]
      ]
  \end{forest}
  \end{minipage}%
  \begin{minipage}{0.5\linewidth}
  \begin{forest}
    for tree = {circle, draw, minimum size = 0.8cm}
      [{$36x$}, fill = julia1
        [{$10x$}, fill = julia1
          [{$3x$}, fill = julia1
            [{$1x$}, fill = julia1]
            [{$2x$}, fill = julia1]
          ]
          [{$7x$}, fill = julia1
            [{$3x$}, fill = julia1] 
            [{$4x$}, fill = julia1] 
          ]
        ]
        [{$26x$}, fill = julia2
          [{$11x$}, fill = julia2
            [{$5x$}, fill = julia2] 
            [{$6x$}, fill = julia2] 
          ]
          [{$15x$}, fill = julia2
            [{$7x$}, fill = julia2] 
            [{$8x$}, fill = julia2] 
          ]
        ]
      ]
  \end{forest}
  \end{minipage}
  \caption{Addition of eight floating point numbers across eight (\textbf{left}) 
  and two (\textbf{right}) machines with a guarantee on SPI. Each machine is 
  represented by a different colour. In both cases, the final result can be 
  represented by the same expression $((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x))$.}
  \label{fig:reduction_tree_colour}
\end{figure*}

### 2025-05-07 workshop/abstract.tex
almost 4 pages but needs work

\begin{abstract}
We introduce a software package, Pigeons.jl, that provides a way to leverage distributed computation to 
obtain samples from complicated probability distributions, such as 
multimodal posteriors arising in Bayesian inference and high-dimensional 
distributions in statistical mechanics. Pigeons.jl provides a simple user interface to 
perform such computations single-threaded, multi-threaded,
and/or distributed over thousands of MPI-communicating machines.
In addition, Pigeons.jl guarantees a property that we call strong parallelism invariance: the 
output for a given seed is identical irrespective of the number of threads and processes, 
which is crucial for scientific reproducibility and software validation. 
We describe the key features of Pigeons.jl and the approach taken to implement 
a distributed and randomized algorithm that satisfies strong parallelism invariance.
\end{abstract}

>>>>

\begin{abstract}
Sampling from complicated probability distributions---such as 
multimodal posteriors arising in Bayesian inference and high-dimensional 
distributions in statistical mechanics---is a challenging problem. Pigeons.jl provides 
a simple user interface to perform such computations single-threaded, multi-threaded,
and/or distributed over thousands of MPI-communicating machines. In this paper,
we discuss a feature of Pigeons.jl that we call \emph{strong parallelism invariance} (SPI): 
a guarantee that the output for a given seed is identical irrespective of the number 
of threads and processes. This feature is crucial for scientific reproducibility and 
software validation. We describe the key features of Pigeons.jl that enable a 
distributed and randomized algorithm satisfying SPI.
\end{abstract}

### 2025-05-07 workshop/examples.tex
almost 4 pages but needs work

\begin{figure}[t]
    \centering 
    \includegraphics[width=0.5\textwidth]{../img/coinflip_posterior.pdf}
    \caption{Posterior distribution for the model given by \cref{eq:coinflip} 
    with $n=100,000$ coin flips and $y=50,000$ observed heads, 
    estimated using $2^{17}$ samples from Pigeons.jl. 
    We present the pairwise plot for $p_1$ and $p_2$, as well as the estimated 
    densities of the marginal of the posterior for each of the two parameters.
    Note that because the model is non-identifiable, as we collect more data the 
    posterior distribution concentrates around the curve $p_1 p_2 = 0.5$, instead 
    of a single point, assuming that the true probability of observing heads 
    is 0.5.}
    \label{fig:coinflip_posterior}
\end{figure}

@@

From here, it is straightforward to sample from the density given by 
\cref{eq:coinflip_density} up to a normalizing constant.
We use non-reversible parallel tempering 
\cite{syed2021nrpt,syed2021paths,surjanovic2022vpt,surjanovic2024ergodicity} (PT), Pigeons.jl's 
state-of-the-art sampling algorithm, to sample from the target distribution.
PT comes with several tuning parameters and \cite{syed2021nrpt} describe how 
to select these parameters effectively, which Pigeons.jl implements under the hood.
We also specify that we would like to store 
the obtained samples in memory to be able to produce trace-plots, as well 
as some basic online summary statistics of the target distribution and useful 
diagnostic output by specifying 
\texttt{record = [traces, online, round\_trip, Pigeons.timing\_extrema, Pigeons.allocation\_extrema]}.
It is also possible to leave the \texttt{record} argument empty and reasonable defaults 
will be selected for the user.
The code below runs Pigeons.jl on one machine with one thread.
We use the default values for most settings, however 
we explain later how one can obtain improved performance by setting 
arguments more carefully (see \cref{sec:additional_options}).
\begin{lstlisting}[language = Julia]
using Pigeons
pt = pigeons(
    target = TuringLogPotential(model), 
    record = [
        traces, online, round_trip, 
        Pigeons.timing_extrema, 
        Pigeons.allocation_extrema])
\end{lstlisting}
Note that to convert the Turing model into an appropriate Pigeons.jl target for sampling, 
we pass the model as an argument to \texttt{TuringLogPotential()}.
Once we have stored the PT output in the variable \texttt{pt} we can 
access the results, as described in the following section. 
The standard output after running the above code chunk is displayed in \cref{fig:standard_out}
and explained in the next section. 
For purposes of comparison, we also run a traditional 
(single-chain Markov chain Monte Carlo) method.

\begin{figure*}[t]
    \centering
    \begin{lstlisting}
            #scans    restarts        Λ      time(s)    allc(B)  log(Z₁/Z₀)   min(α)     mean(α)

                  2          0       1.04      0.383   3.48e+07  -4.24e+03          0      0.885
                  4          0       4.06    0.00287   1.79e+06      -16.3   4.63e-06      0.549
                  8          0       3.49    0.00622   3.55e+06      -12.1      0.215      0.612
                 16          0       2.68     0.0161   7.46e+06      -10.2      0.518      0.703
                 32          0       4.29     0.0353   1.37e+07      -11.8      0.222      0.524
                 64          3       3.17     0.0699   2.86e+07      -11.5      0.529      0.648
                128          8       3.56      0.139   5.53e+07      -11.5      0.523      0.605
                256         12       3.38      0.241    1.1e+08      -11.6      0.526      0.625
                512         37       3.48      0.473   2.22e+08        -12      0.527      0.614
           1.02e+03         77       3.55      0.895   4.46e+08      -11.8      0.571      0.605

    \end{lstlisting}
    \caption{Standard output provided by Pigeons.jl. 
    Rows indicate tuning rounds of the PT algorithm with an exponentially increasing number 
    of PT iterations (\texttt{\#scans}). Columns indicate various useful diagnostics, 
    such as the amount of memory allocations per round (in bytes), time (in seconds), and estimates 
    of the log of the normalization constants. 
    The output is described in greater detail in \cref{sec:PT_diagnostics}
    and has been modified to exclude columns that are not described in the paper.}
    \label{fig:standard_out}
\end{figure*}

@@

\subsection{Outputs}
Pigeons.jl provides many useful types of output, such as: plots of samples from 
the distribution, estimates of normalization constants, summary statistics of 
the target distribution, and various other diagnostics. 
We describe several examples of possible output below.

\subsubsection{Standard output}
An example of the standard output provided by Pigeons.jl is displayed in \cref{fig:standard_out}. 
Each row of the table in the output indicates a new tuning round in parallel tempering, 
with the \texttt{\#scans} column indicating the number of scans/samples in 
that tuning round. 
During these tuning rounds, Pigeons.jl searches for optimal values of certain PT 
tuning parameters. Other outputs include: 
\begin{itemize}
    \item \texttt{restarts}: a higher number is better. 
    Informally, PT propagates samples from an easy-to-sample distribution (the reference) 
    to the more difficult target distribution. 
    A tempered restart happens when a sample from the reference successfully 
    percolates to the target. (See the subsequent sections for a more detailed description 
    of parallel tempering.) 
    When the reference supports i.i.d.~sampling, tempered restarts 
    can enable large jumps in the state space.

    \item \texttt{$\Lambda$}: the global communication barrier, as described in \cite{syed2021nrpt}, 
    which measures the inherent difficulty of the sampling problem.
    A rule of thumb to configure the number of PT chains is also 
    given by \cite{syed2021nrpt}, where they suggest that stable performance 
    should be achieved when the number of chains is set to roughly 2$\Lambda$.
    See \cref{sec:PT_diagnostics} for more information.

    \item \texttt{time} and \texttt{allc}: the time (in seconds) and number of 
    allocations (in bytes) used in each round.

    \item \texttt{$\log(Z_1/Z_0)$}: an estimate of the 
    logarithm of the ratio of normalization constants between the target and the reference. 
    In many cases, $Z_0 = 1$.

    \item \texttt{min($\alpha$)} and \texttt{mean($\alpha$)}: minimum and average swap 
    acceptance rates during the communication phase across the PT chains.
    See \cref{sec:PT} for a description of PT communication.
\end{itemize}

\subsubsection{Plots}
It is straightforward to obtain plots of samples from the target distribution, 
such as trace-plots, pairwise plots, and density plots of the marginals.

To obtain posterior densities and trace-plots, we first 
make sure that we have the third-party 
MCMCChains.jl\footnote{\url{https://github.com/TuringLang/MCMCChains.jl}}, 
StatsPlots.jl\footnote{\url{https://github.com/JuliaPlots/StatsPlots.jl}}, and 
PlotlyJS.jl\footnote{\url{https://github.com/JuliaPlots/PlotlyJS.jl}} 
packages installed via
\begin{lstlisting}[language=Julia]
Pkg.add("MCMCChains", "StatsPlots", "PlotlyJS")
\end{lstlisting}

With the \texttt{pt} output object from before for our non-identifiable coinflip model, 
we can run the following:
\begin{lstlisting}[language=Julia]
using MCMCChains, StatsPlots, PlotlyJS
plotlyjs()
samples = Chains(
    sample_array(pt), variable_names(pt))
my_plot = StatsPlots.plot(samples)
display(my_plot)
\end{lstlisting}
The output of the above code chunk is an interactive plot that can be zoomed in or out
and exported as an HTML webpage. 
A modified static version of the output for the first parameter, $p_1$, 
is displayed in the top panel of \cref{fig:trace_density_plots},
along with a comparison to the output from a single-chain algorithm in the bottom 
panel of the same figure.

\begin{figure*}[t]
    \centering
    \begin{minipage}{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/trace_density_plots_pt.pdf}
    \end{minipage}
    \begin{minipage}{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/trace_density_plots_single_chain.pdf}
    \end{minipage}
    \caption{
        Trace-plots for the first parameter, $p_1$, in the 
        non-identifiable coinflip Turing model.
        \textbf{Left:} Samples from Pigeons.jl using PT with 10 chains. Note that the 
        trace-plot indicates fast mixing/exploration across the state space.
        \textbf{Right:} Single-chain Markov chain Monte Carlo. Note that the 
        trace-plot explores the state space much more slowly when we do not use PT.
    }
    \label{fig:trace_density_plots}
\end{figure*}

To obtain pair plots, we add the 
PairPlots.jl\footnote{\url{https://github.com/sefffal/PairPlots.jl}}
and CairoMakie.jl\footnote{\url{https://github.com/JuliaPlots/CairoMakie.jl}}
packages:
\begin{lstlisting}[language=Julia]
Pkg.add("PairPlots", "CairoMakie")
\end{lstlisting}
and then run
\begin{lstlisting}[language=Julia]
using PairPlots, CairoMakie
my_plot = PairPlots.pairplot(samples) 
display(my_plot)
\end{lstlisting}
The output of the above code chunk with an increased number of samples 
is displayed in \cref{fig:coinflip_posterior}.
\footnote{The code chunk above requires Julia 1.9.}

\subsubsection{Estimate of normalization constant}
The (typically unknown) constant $Z$ in \cref{eq:normalizing_constant} is 
referred to as the \emph{normalization constant}. 
In many applications, it is useful to approximate this constant. 
For example, in Bayesian statistics, this corresponds to the 
marginal likelihood and can be used for model selection.

As a side-product of PT, we automatically obtain an approximation to the natural 
logarithm of the normalization constant. This is done automatically using the 
stepping stone estimator \cite{xie2011improving}.
The estimate can be accessed using
\begin{lstlisting}[language=Julia]
stepping_stone(pt)
\end{lstlisting}
In the case of the normalization constant given by \cref{eq:coinflip_normalization} 
for $n=100,000$ and $y=50,000$, we can exactly obtain its value as $\log(Z) \approx -11.8794$.
Note that this is very close to the output provided in \cref{fig:standard_out}.

\subsubsection{Online statistics}
\label{sec:online_stats}
Pigeons has facilities to support cases requiring large memory. For instance, we
allow for the computation of online statistics, as well as off-memory sample storage.
Having specified the use of the \texttt{online} recorder in our call to \texttt{pigeons()}, 
we can output some basic summary statistics of the marginals of our target distribution. 
For instance, it is straightforward to estimate the mean and variance of each 
of the marginals of the target with 
\begin{lstlisting}[language=Julia]
using Statistics
mean(pt); var(pt)
\end{lstlisting}
Other constant-memory statistic accumulators are made available in the 
OnlineStats.jl \cite{day2020onlinestats} package. 
To add additional constant-memory statistic accumulators, we 
can register them via \texttt{Pigeons.register\_online\_type()}, as described in our 
online documentation. For instance, we can also compute constant-memory estimates 
of extrema of our distribution.

>>>>

\begin{figure}[t]
    \centering 
    \includegraphics[width=0.8\linewidth]{../img/coinflip_posterior.pdf}
    \caption{Posterior distribution for the model given by \cref{eq:coinflip} 
    with $n=100,000$ coin flips and $y=50,000$ observed heads, 
    estimated using $2^{17}$ samples from Pigeons.jl. 
    We present the pairwise plot for $p_1$ and $p_2$, as well as the estimated 
    densities of the marginal of the posterior for each of the two parameters.
    Note that because the model is non-identifiable, as we collect more data the 
    posterior distribution concentrates around the curve $p_1 p_2 = 0.5$, instead 
    of a single point, assuming that the true probability of observing heads 
    is 0.5.}
    \label{fig:coinflip_posterior}
\end{figure}

@@

From here, it is straightforward to sample from the density given by 
\cref{eq:coinflip_density} up to a normalizing constant.
We use non-reversible parallel tempering 
\cite{syed2021nrpt,syed2021paths,surjanovic2022vpt,surjanovic2024ergodicity} (PT), Pigeons.jl's 
state-of-the-art sampling algorithm, to sample from the target distribution.
PT comes with several tuning parameters and \cite{syed2021nrpt} describe how 
to select these parameters effectively, which Pigeons.jl implements under the hood.
We also specify that we would like to store 
the obtained samples in memory to be able to produce trace-plots, as well 
as some basic online summary statistics of the target distribution and useful 
diagnostic output by specifying 
\texttt{record = [traces, online, round\_trip, Pigeons.timing\_extrema, Pigeons.allocation\_extrema]}.
It is also possible to leave the \texttt{record} argument empty and reasonable defaults 
will be selected for the user.
The code below runs Pigeons.jl on one machine with one thread.
We use the default values for most settings, however 
we explain later how one can obtain improved performance by setting 
arguments more carefully (see \cref{sec:additional_options}).
\begin{lstlisting}[language = Julia]
using Pigeons
pt = pigeons(
    target = TuringLogPotential(model), 
    record = [
        traces, online, round_trip, 
        Pigeons.timing_extrema, 
        Pigeons.allocation_extrema])
\end{lstlisting}
Note that to convert the Turing model into an appropriate Pigeons.jl target for sampling, 
we pass the model as an argument to \texttt{TuringLogPotential()}.
Once we have stored the PT output in the variable \texttt{pt} we can 
access the results, as described in the following section. 
For purposes of comparison, we also run a traditional 
(single-chain Markov chain Monte Carlo) method.

@@

### 2025-05-07 workshop/introduction.tex
almost 4 pages but needs work

Pigeons.jl\footnote{The source code for Pigeons.jl can be found at 
\url{https://github.com/Julia-Tempering/Pigeons.jl}. 
Pigeons.jl v0.2.0, used in the example scripts below, is compatible with Julia 1.8+.}
enables users to sample efficiently from challenging probability distributions 
and solve integration problems by 
implementing state-of-the-art sampling algorithms \cite{syed2021nrpt,surjanovic2022vpt} 
that leverage distributed computation. Its simple user interface facilitates performing such computations 
single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
We have stress-tested Pigeons successfully with up to 1000 machines running 
concurrently on a compute cluster (Sockeye) at the University of British Columbia.
Further, Pigeons comes with guarantees on strong parallelism invariance wherein 
the output for a given seed is \emph{identical} irrespective 
of the number of threads or processes. 
Such a level of reproducibility is rare in distributed software but of 
great use for the purposes of debugging in the context of sampling algorithms, 
which produce stochastic output.
Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
\begin{enumerate}
    \item one machine running on one thread;
    \item one machine running on several threads;
    \item several machines running, each using one thread, and
    \item several machines running, each using several threads.
\end{enumerate}

\subsection{Latest documentation}

More in-depth documentation and additional examples can be found at: {\renewcommand{\UrlFont}{\ttfamily\normalsize}\url{https://pigeons.run/}}.

\subsection{Problem formulation}
We describe the class of problems that can be approached using Pigeons.jl.
Let $\pi(x)$ denote a probability density\footnote{This density may 
also be a probability mass function 
(discrete variables). We also allow for a combination of discrete/continuous variables.} 
called the \emph{target}. 
In many problems, e.g. in Bayesian statistics, the density $\pi$ is typically 
known only up to a normalizing constant, 
\[
\label{eq:normalizing_constant}
  \pi(x) = \frac{\gamma(x)}{Z}, \qquad Z = \int \gamma(x) \, \dee x,
\]  
where $\gamma$ can be evaluated pointwise but $Z$ is typically unknown.
Pigeons.jl takes as input the function $\gamma$.

@@

\subsection{What is of interest to the general Julia developer?}
Ensuring code correctness at the intersection of randomized, parallel, and 
distributed algorithms is a challenge. To address this challenge, we designed Pigeons.jl 
based on a principle that we refer to as \textit{strong parallelism invariance}
(SPI).
Namely, the output of Pigeons.jl is  invariant to the number of machines 
and/or threads.
Without explicitly keeping SPI in mind during software construction, 
the (random) output of the algorithm is only guaranteed to have the same distribution.
This is a much weaker guarantee that, in particular, makes debugging difficult.
However, with our notion of SPI we make debugging and 
software validation considerably easier. This is because the developer can 
first focus on the fully serial randomized algorithm, and then use it as an 
easy-to-compare gold-standard reference for parallel/distributed implementations. 
This strategy is used extensively in Pigeons.jl to ensure correctness. In contrast, 
testing equality in distribution, while possible (e.g., see \cite{geweke2004getting}), 
incurs additional false negatives due to statistical error.

The general Julia developer will be interested in: 
\begin{itemize}
    \item The main causes of a violation of strong parallelism invariance that we have identified 
    (\cref{sec:PI_causes})---some of which are specific to Julia---and how we address 
    them in Pigeons.jl.

    \item The SplittableRandoms.jl\footnote{\url{https://github.com/Julia-Tempering/SplittableRandoms.jl}}
    package that was developed by our team 
    to achieve strong parallelism invariance in Pigeons.jl (\cref{sec:splittable_randoms}).
\end{itemize}

>>>>

Pigeons.jl\footnote{The source code for Pigeons.jl can be found at 
\url{https://github.com/Julia-Tempering/Pigeons.jl}. 
Pigeons.jl v0.2.0, used in the example scripts below, is compatible with Julia 1.8+.}
enables users to sample efficiently from challenging probability distributions 
and solve integration problems by 
implementing state-of-the-art sampling algorithms \cite{syed2021nrpt,surjanovic2022vpt} 
that leverage distributed computation. Its simple user interface facilitates performing such computations 
single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
We have stress-tested Pigeons successfully with up to 1000 machines running 
concurrently on a compute cluster (Sockeye) at the University of British Columbia.
Importantly, Pigeons comes with guarantees on \emph{strong parallelism invariance} (SPI), 
wherein the output for a given seed is \emph{identical} irrespective of the number 
of threads or processes. Such a level of reproducibility is rare in distributed 
software but of great use for the purposes of debugging in the context of sampling 
algorithms, which produce stochastic output.
Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
\begin{enumerate}
    \item one machine running on one thread;
    \item one machine running on several threads;
    \item several machines running, each using one thread, and
    \item several machines running, each using several threads.
\end{enumerate}

\section{Overview of Pigeons.jl}
A typical instance of the class of problems that Pigeons.jl tackles can be described by
a probability density\footnote{This density may also be a probability mass function 
(discrete variables). We also allow for a combination of discrete/continuous variables.} 
$\pi(x)$ called the \emph{target}. In many problems, e.g. in Bayesian statistics, the 
density $\pi$ is typically known only up to a normalizing constant, 
\[
\label{eq:normalizing_constant}
  \pi(x) = \frac{\gamma(x)}{Z}, \qquad Z = \int \gamma(x) \, \dee x,
\]  
where $\gamma$ can be evaluated pointwise but $Z$ is typically unknown.
Pigeons.jl takes as input the function $\gamma$.

@@

### 2025-05-07 workshop/paper.tex
some fixes
almost 4 pages but needs work

\usepackage[accepted]{icml2025}
\usepackage{listings}
\usepackage{jlcode}
\usepackage{natbib}
\setcounter{page}{1}

\usepackage{expl3} % add missing functionality in the old texlive version running on the server 
\usepackage[autonum, blackhypersetup]{shortex}
\usepackage{xcolor}
%\let\subcaption\undefined % overwrite 'subcaption' from the style file
%\usepackage[font=scriptsize]{subcaption}
\usepackage{forest}

\newcommand{\NS}[1]{\textcolor{red}{#1}}
\def\bfx{\mathbf{x}}
\def\bfX{\mathbf{X}}

\definecolor{julia1}{rgb}{0, 0.60, 0.98}
\definecolor{julia2}{rgb}{0.89, 0.43, 0.28}
\definecolor{julia3}{rgb}{0.24, 0.64, 0.30}
\definecolor{julia4}{rgb}{0.76, 0.44, 0.82}
\definecolor{julia5}{rgb}{0.67, 0.55, 0.09}
\definecolor{julia6}{HTML}{00aaae}
\definecolor{julia7}{HTML}{ed5e93}
\definecolor{julia8}{HTML}{c68225}

\pdfsuppresswarningpagegroup=1 % avoid warning due to pdfs made with photoshop

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Pigeons.jl: Distributed sampling from intractable distributions}

@@

\twocolumn[
\icmltitle{Pigeons.jl: Distributed sampling from intractable distributions}

@@

\input{abstract}
\input{introduction.tex}
\input{examples.tex}
\input{PT.tex}
\input{SPI.tex}
\input{related.tex}
\input{conclusion.tex}
\input{ack}

@@

>>>>

\input{macros}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\newcommand{\thetitle}{Reproducible sampling from intractable distributions with Pigeons.jl}
\icmltitlerunning{\thetitle}

@@

\twocolumn[
\icmltitle{\thetitle}

@@

\input{abstract}
\input{introduction.tex}
\input{SPI.tex}
\input{related.tex}
%\input{conclusion.tex}
%\input{ack}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn

\input{examples}
\input{PT}
